Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-04-06 manuscript/manuscript.qmd
Revised introduction

---
title: "Flattering Advice: Avoiding Disappointment in Advice-Giving"
shorttitle: "Flattering Advice"
date: yes
author: 
  - name: "Amanda Chen"
    corresponding: true
    orcid: 0000-0002-4428-627X
    email: zchengj@connect.ust.hk
    affiliations:
      - id: id1
        name: "The Hong Kong University of Science and Technology"
        department: "Department of Management"
        city: "Hong Kong"
  - name: "David Hagmann"
    orcid: 0000-0002-2080-997X
    affiliations:
    - id: id1
# author-note:
  # disclosures:
    # gratitude: The authors are grateful to XXX for their helpful comments.
abstract: Advice can improve decisions when it offers an unbiased evaluation of information, but the painful truth can disappoint an advisee's overoptimistic expectations. Across three pre-registered and incentivized experiments involving real adviser-advisee interactions (N = 3,900), we show that advisers avoid such disappointment at the expense of accuracy and their own earnings. Participants inflate the attractiveness of advisees and do so to a greater extent when incentivized to be viewed as likeable. Advisers who provide flattering advice are rated as more likable and trustworthy, even as their advice is less honest and accurate (Study 1). When advisers observe an advisee's quiz performance along with a signal expressing a preference for competing against a group of low or higher performers, the (uninformative) signal influences their advice (Study 2). Finally, in a context where men expect to do better than women, advisers who observe expectations (but not gender) on top of actual performance are more likely to tell men to enter a tougher (but potentially more rewarding) competition. Advisers' efforts to avoid disappointment may thus contribute to systematic gender disparities in advice.

@@

Advice figures prominently in workplace and personal decision-making,
shaping choices about promotions, career paths, and everyday dilemmas
[@Gino2007; @Harvey1997; @Soll2009]. While others can provide
information in the form of opinions [@Trouche2018], feedback
[@Kluger1996], or personal observations [@Yoon2021] to influence the
decision-maker, this project focuses specifically on advice—defined as
"information that is perceived as intended to help an advisee deal with
their issue of deliberation" [@Milyavsky2024]. Advice is distinct from
other forms of input because its primary purpose is to improve
decision-making by leveraging others' expertise.

Advice has the power to enhance decisions, yet it also carries the risk
of disappointment. For example, a career consultant might advise an
employee to enroll in a management program to strengthen their skill set
before competing for a leadership role. Ideally, such honest advice
helps the employee identify areas for personal growth. In practice,
however, the employee might feel overlooked and frustrated that their
ambitions are not immediately supported. As a result, advisers may be
inclined to offer flattering recommendations that align with—or even
exceed—advisees' hopes, such as encouraging the employee to pursue the
leadership role despite concerns about their readiness. In this way,
interpersonal dynamics may discourage full candor in giving advice.

@@

# Overview of studies

@@

# Open Science Statement

@@

# Study 1

@@

## Methods

@@

## Results

@@

## Discussion

@@

# Study 2

@@

# Study 3

@@

# General Discussion

>>>>

---
title: "Flattering Advice: Avoiding Disappointment in Advice-Giving"
shorttitle: "Flattering Advice"
date: yes
author: 
  - name: "Amanda Chen"
    corresponding: true
    orcid: 0000-0002-4428-627X
    email: zchengj@connect.ust.hk
    affiliations:
      - id: id1
        name: "The Hong Kong University of Science and Technology"
        department: "Department of Management"
        city: "Hong Kong"
  - name: "David Hagmann"
    orcid: 0000-0002-2080-997X
    affiliations:
    - id: id1
# author-note:
  # disclosures:
    # gratitude: The authors are grateful to XXX for their helpful comments.
abstract: Advice can improve decisions when it offers an unbiased evaluation of information, but the painful truth can disappoint an advisee's expectations. Across three pre-registered and incentivized experiments involving real adviser-advisee interactions (N = 3,900), we show that advisers avoid delivering such disappointment at the expense of accuracy and their own earnings. Participants inflate the attractiveness of advisees and do so to a greater extent when incentivized to be viewed as likeable. Advisers who provide flattering advice are indeed rated as more likable and trustworthy, even as their advice is less honest and accurate (Study 1). When advisers observe an advisee's quiz performance along with a signal expressing a preference for competing against a group of low or higher performers, the (uninformative) signal influences the advice they give (Study 2). Finally, in a context where men expect to do better than women, advisers who observe expectations (but not gender) on top of actual performance are more likely to tell men to enter a tougher (but potentially more rewarding) competition. Advisers' efforts to avoid disappointment may thus contribute to systematic gender disparities in advice and outcomes.

@@

Advice figures prominently in professional and personal life,
shaping choices about career paths, promotions, and everyday dilemmas
[@Gino2007; @Harvey1997; @Soll2009].
An extensive body of research has examined the behavior and beliefs of those
receiving advice, including how willing they are to seek advice [@Heikensten2019], 
whom they want to get advice from [@Hur2020], and how much they revise their beliefs 
in response to advice [@Bailey2023].
In this paper, we shift the focus to the behavior of advisers. Prior work 
assumes that they (absent conflicts of interest) are motivated 
to help someone else make a better decision, even defining advice as being 
intended to help another decision-maker [@Milyavsky2024 -- David Q: is this a right interpretation of the previous quote here?].
We propose that advisers are not merely concerned about the outcome of the action
they recommend, however, but also consider the hedonic impact of the advice itself.
Across three experiments, we show that advisers are responsive to the expectations
of an advisee and tailor their recommendations to avoid disappointing them,
even when doing so leads to worse advice and is costly for both parties.

In many situations, people turn to those who are better informed to advise them
on what actions to take. For example, an employee may discuss career
opportunities with a supervisor or mentor to see if they should apply for a
promotion or take on a risky project. Ideally, they would receive an unbiased
assessment based on the adviser's experience and impression of the advice-seeker,
taking into account the potential costs (e.g., opportunity costs) and benefits
(e.g., promotion chances and compensation). Indeed, in an experimental context, 
advisers are helpful in encouraging
high-performers to enter competitions and discouraging low-performers from doing so
[@Brandts2015]. However, people often reveal their expectations along with a
request for advice. For example, someone may express a desire to take on more
challenging projects before asking for advice on whether to do so. In such cases,
honest advice may discourage someone from an action they intended to take,
frustrating and disappointing them in the process. An adviser who is concerned
about the hedonic impact of information on the recipient, for example because
they recognize the psychological cost of disappointment [@Gul1991; @Golman2017]
or do not want to be blamed for delivering bad news [@John2019], 
may give advice that makes the recipient feel good in the moment, even if they 
expect it to lead to a bad outcome. We refer to such advice as "flattering"
because it communicates a more favorable belief about the recipient than the
sender in fact holds.

In the workplace, prior research has documented stark gender differences in what
kind of advice employees receive. In particular, men receive more critical 
performance feedback than women, which although
painful in the moment, facilitates learning and leads to better performance
in the future (CITE). While such differences
are often attributed to unconscious biases, we propose that avoiding disappointment
may be another, belief-based contributing factor. Prior work has found that
men are more confident in their performance than equally well-performing
women [@Exley2024; @Lundeberg1994; @Mobius2022]. As a result,
women are less likely to enter into competitions and less likely to apply for 
more challenging (but better compensated) work [@Niederle2007; @Coffman2024].
We propose that in addition to self-selection, confidence may also impact the
advice men and women receive. An adviser who is motivated to avoid disappointment
will take into account how someone *expects* to perform, even when they are
fully informed about how that individual *actually* performs. Thus, people
who are more (over)confident receive more flattering advice, leading to gender
differences driven by differences in expectations.

We present results from three preregistered and incentivized experiments involving
advisers and advisees. Each study consists of three asynchronous stages with a group of
participants performing an initial task, a second group of participants providing
advice, and the original participants returning to receive the advice and make
a decision. Advisers have a financial incentive for the advisee to make a good
decision, except in one treatment in Study 1 where they are incentivized to be
viewed as likeable.

In the first study, participants upload photos of themselves and
are matched with nine other participants of the same gender. A new group of
participants of the opposite gender then rank the selfies and provide advice
to one participant (whom they ranked low on attractiveness) on what rank they
should bet on. The original participants return, receive advice, bet on
how attractive other participants ranked them, and evaluate the adviser on
measures such as likeability and trustworthiness. Advisers who receive an
incentive for being rated as likeable advise the other participant to bet that
they are more attractive, relative to those incentivized for accuracy and relative
to their own assessment. Advisees, moreover, evaluate those who flatter them as
more likeable and more trustworthy, even as they receive less honest and less
accurate advice.

In Study 2, participants complete a quiz assessing different personally relevant
skills (problem-solving,
emotional intelligence, and communications) and make a non-binding decision about
whether they would like to compete against a group that did well on this quiz
or a group that did poorly. Competing against high performers offers the opportunity
of a higher payment, but requires a higher score. We randomly assign participants
to have one of the two groups selected by default. Advisers then observe a
participant's real score on the quiz and, in one treatment, also which of the
two groups they had selected. We show that participants' random assignment to
a default influences the advice they receive when their choice (but not the
random assignment) is visible.

Finally, we look at a setting in which we expect gender differences in confidence.
Participants complete a mathematics quiz and make an (unincentivized) guess about
their score. Advisers then observe the actual quiz score and, in one treatment,
also the unincentivized guess. They then advise the participant whether to
compete against a group of high performers or a group of low performers. As 
expected, men are more confident in their performance than are women, holding
fixed their actual performance. Importantly, we show that when participants' 
guesses are shown to advisers, men are more likely than women to be advised to
compete against the strong group.

The remainder of this paper is organized as follows. Section 2 provides an
overview of the literature on advice and disappointment aversion. Section 3
presents results from three preregistered and incentivized experiments
involving adviser-advisee pairs. Section 4 concludes.

# Literature Review

@@

## Disappointment Aversion
Information, including advice, has traditionally been seen as valuable when 
(and to the extent that) it leads to better decisions [@Stigler1961]. However, 
even information that is available at no economic cost can be unpleasant and 
thus impose a psychological cost [@Golman2017; @Ho2021; @Loewenstein2018]. One reason
for why information can be painful to receive is because it leads to disappointment
[@Gul1991]. Disappointment is costly if it suggests that a desired outcome is
not realized, but also because it provides information about future consumption
[@Koszegi2010]. Someone receiving advice against applying for a promotion, for
example, would not only learn that they are not as competitive as they had
believed, but also has to update their beliefs about their future income stream.

@Bell1985 shows that disappointment aversion implies providing good news
immediately, but breaking bad news piecewise to adjust expectations---and 
observes that this is indeed how people inform others. Notably, this is assuming
delivering the bad news is unavoidable. In the context of advice, however,
people can elect not to deliver the bad news at all and instead shade their
evaluation to meet expectations and flatter the recipient.

## Gender Differences in Advice and Expectations

[A few sentences summarizing the literature review and tying it closely to the experiments.]

# Experiments

@@

## Open Science Statement

@@

## Study 1

@@

### Methods

@@

### Results

@@

### Discussion

@@

## Study 2

@@

## Study 3

@@

# Discussion
